---
title: "New Data Prep"
author: "Miles D. Williams"
date: 'Created: 2023-07-19 | Modified `r Sys.Date()`'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = F,
                      warning = F)
```

## Setup

I'll be using the `{peacesciencer}` package to access the data I need. The below code chunk opens the package, in addition to tools in the `{tidyverse}` for data wrangling.

```{r}
library(tidyverse)
library(peacesciencer)
```

## Data Wrangling

### Country-Year-Conflict Data

First, create a directed dyad dataset and add COW data on major international wars. There should be 95 unique conflicts in the dataset.

```{r}
ddy_dt <- create_dyadyears() |>
  ## add cow war data
  left_join(
    cow_war_inter
  )

## check that no. of unique wars = 95
length(na.omit(unique(ddy_dt$warnum))) == 95
```

Next, I need to convert the data to country-year and then merge in additional economic, polity, and militarization data.

```{r}
cy_dt <- ddy_dt |>
  ## select the columns I need
  select(year, ccode1, warnum, cowinteronset, batdeath1) |>
  ## only keep observations for years that a war starts
  filter(cowinteronset == 1) |>
  ## don't need the onset indicator anymore
  select(-cowinteronset) |>
  ## rename
  rename(ccode = ccode1, batdeath = batdeath1)
```

Now, I need to add in covariates.

```{r}
cy_dt <- cy_dt |>
  ## economic data (surplus domestic product)
  left_join(cow_sdp_gdp) |>
  ## military capabilities data
  left_join(cow_nmc) |>
  ## democracy data
  left_join(ccode_democracy)
cy_dt

## To learn more about these datasets:
#?cow_sdp_gdp
#?cow_nmc
#?ccode_democracy
```

Okay, now there should be some duplicates that arise for wars that have more than two countries involved. I need to drop these.

```{r}
orig_n <- nrow(cy_dt)
cy_dt_nodups <- cy_dt |>
  distinct()
nodup_n <- nrow(cy_dt_nodups)
c(original = orig_n, nodups = nodup_n)
```


Create some additional variables.

```{r}
cy_dt_nodups |>
  mutate(
    ## normalize battle deaths by country population
    batdeathpc = batdeath / (tpop * 1000),
    ## normalize military expenditures by surplus domestic product
    milexpsdp = (milex * 1000) / exp(sdpest)
  ) -> cy_dt_nodups
```

Save as a .csv file:

```{r}
write_csv(
  cy_dt_nodups,
  here::here("01_data", "country-year-war.csv")
)
```


### Conflict-Year Data

Now, let's create a version of the data aggregated to the level of particular wars.

```{r}
cy_dt_nodups |>
  group_by(warnum) |>
  summarize(
    year = min(year),
    batdeath = sum(batdeath),
    batdeathpc = batdeath / sum(tpop * 1000),
    ## sums
    across(
      c(tpop, milper, milex),
      ~ sum(.x, na.rm=T)
    ),
    across(
      c(sdpest, wbgdp2011est),
      .fns = list(
        "sum" = ~ sum(exp(.x), na.rm=T),
        "mea" = ~ mean(exp(.x), na.rm=T),
        "med" = ~ median(exp(.x), na.rm=T),
        "min" = ~ min(exp(.x), na.rm=T),
        "max" = ~ max(exp(.x), na.rm=T)
      ),
      .names = "{.fn}_{.col}"
    ),
    min_polyarchy = min(v2x_polyarchy, na.rm=T),
    med_polyarchy = median(v2x_polyarchy, na.rm=T),
    mea_polyarchy = mean(v2x_polyarchy, na.rm=T),
    max_polyarchy = max(v2x_polyarchy, na.rm=T),
    min_polity2 = min(polity2, na.rm=T),
    min_polity2 = ifelse(min_polity2==Inf, NA, min_polity2),
    med_polity2 = median(polity2, na.rm=T),
    mea_polity2 = mean(polity2, na.rm=T),
    max_polity2 = max(polity2, na.rm=T),
    max_polity2 = ifelse(max_polity2==-Inf, NA, max_polity2)
  ) -> wy_dt
```

Add total population by year:

```{r}
cow_nmc |>
  group_by(year) |>
  summarize(
    wpop = sum(tpop * 1000, na.rm=T)
  ) -> world_pop
wy_dt <- left_join(wy_dt, world_pop)
```

```{r}
write_csv(
  wy_dt,
  here::here("01_data", "war-year.csv")
)
```

