---
# title: "A Logistic Model of War Deaths is Better than the Power-Law"
# author: "Miles D. Williams"
# date: '2022-12-12'
bibliography: ../05_ref/ref.bib
editor_options: 
  markdown: 
    wrap: 72
fontfamily: mathpazo
fontsize: 12pt
geometry: margin=1in
header-includes:
- \usepackage{float}
- \usepackage{setspace}
- \usepackage{dcolumn}
- \doublespacing
- \setlength\parindent{24pt}
- \setlength\parskip{1pt}
- \usepackage[group-separator={,}]{siunitx}
- \usepackage{threeparttable}
- \usepackage{caption}\DeclareCaptionLabelFormat{AppendixTables}{Table A.#2}
- \usepackage{caption}\DeclareCaptionLabelFormat{AppendixFigures}{Figure A.#2}
- \usepackage[labelsep=period]{caption}
output:
  pdf_document:
    number_sections: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      message = F,
                      warning = F,
                      fig.align = "center")
library(knitr)
library(here)
```

## Introduction

In the social sciences, few regularities attain law-like status. The
idea that war's deadliness follows a power-law distribution holds this
coveted spot. The power-law is a simple heavy-tailed probability
distribution used to describe a range of phenomena that have many small
values and only a few large ones. Lewis F. Richardson first noticed that
international war fatalities follow such a distribution in the
mid-twentieth century [@richardson1948; @richardson1960], and in the
years since his initial discovery scholars attributed the power-law to
various other dimensions of war [@cioffirevillaEtAl2010] and to several
other types of violence, including terrorism and civil wars
[@clauset2020]. This places human conflict in the company of
wide-ranging phenomena, from earthquakes to solar flares, and fluid
turbulence to city population size [@clausetEtAl2009].

While the idea that wars obey the power-law has a long tradition in IR,
it received special attention with the publication of Bear Braumoeller's
[-@braumoeller2019] book *Only the Dead*, which provides a thorough,
data-driven rebuttal to claims made by proponents of the so-called
decline-of-war thesis [see @pinker2011]. Adherents to the view claim
that for reasons ranging from Enlightenment Humanism to shifting
cultural norms, human violence is in a secular decline and has been for
some time.

Contrary to this optimistic view of human nature, @braumoeller2019
finds, among many things, that power-law fits to international wars
before and after 1945 (a significant pivot-point for decline-of-war
theorists) are statistically indistinguishable and each consistent with
a potentially infinite mean for war fatalities---a unique possibility
that can arise in power-law probability distributions. Taken together,
these findings simultaneously deal a blow to a key claim of the
decline-of-war thesis and warn against complacency about the deadly
potential of any future war.

While not questioning the totality of evidence @braumoeller2019 levies
against the decline-of-war thesis, results from this study raise
questions specifically about the power-law's applicability to war
intensity. Does an alternative probability distribution provide a better
fit for the data? If so, does this alternative provide different
inferences about whether a shift in war's deadliness occurred in the
last century?

The case is made here that the power-law *does not* provide the best fit
for the data. Rather, a log-logistic distribution provides a better fit
and supports different inferences about a change in war intensity. Using
data on the intensity of major international wars from the past 200
years, a comparison of the power-law model and a log-logistic model
demonstrates the superiority of the latter. The log-logistic model of
war intensity not only provides a much better fit for *all* of the data,
its predictions have far greater efficiency than that of the power-law
in the extreme tails of the distribution of war intensity.

This finding has important general implications for theory, statistical
inference, and hypothesis testing with war intensity data [see
@cederman2003; @cedermanEtAl2011; and @friedman2014]. More specifically,
as further replication and extension of the analysis done by
@braumoeller2019 shows, the choice between using the power-law or the
log-logistic probability distribution leads to different conclusions
about whether a significant shift in war's deadliness occurred after
1945. Consistent with Braumoeller's [-@braumoeller2019] findings, when
modeled with the power-law, a statistical test fails to reject the null
that war intensities pre- and post-1945 come from different
distributions. However, when modeled with the log-logistic distribution,
a statistical test rejects the null, and the data are consistent with a
substantial decline in war intensity after 1945.

The paper proceeds as follows. Motivation and background on modeling
wars with the power-law is provided. A summary of the power-law model of
war intensity and an alternative log-logistic specification follows. I
then summarize the data used to compare the power-law and log-logistic
modeling approaches. Data on war fatalities come from the Correlates of
War (COW) Project. War intensity is calculated as the total battle
deaths of all countries involved in a war per the combined population of
the relevant countries at the time of the conflict. Next, results
comparing the power-law with the log-logistic fit are presented,
followed by a replication and extension of @braumoeller2019.

The final section addresses the implications of these findings. While
the power-law may apply in numerous settings, in the case of
international wars it provides a poorer fit than decades of tradition
would lead one to believe. In contrast to the power-law, the
log-logistic model offers a better fit and does so with greater
statistical precision. This finding has significant ramifications for
the quantitative study of war. Scholars have predicated theoretical
models of war's microfoundations, missing data imputation, and
hypothesis testing on the power-law [@cederman2003; @cedermanEtAl2011;
@friedman2014]. A range of important research may need to be reevaluated
to confirm its robustness.

Outside of its implications for quantitative research on war, the theme
of this study falls in line with a broader trend across scientific
fields questioning the appropriateness of the power-law where once
scholars took it for granted. For example, recent studies show that more
than a 1,000 real-world networks [@broidoClauset2019] and even solar
flares [@verbeeckEtAl2019], which researchers long thought obeyed the
power-law, instead may arguably follow a log-normal distribution
(something very close to the log-logistic). The time, therefore, is well
overdue to revisit the power-law and war.

## Power-Laws and War Deaths

Regarding power-laws and war, the title of Johnson and colleagues'
[-@johnsonEtAl2013] article in *Scientific Reports* states plainly that
a "Simple Mathematical Law Benchmarks Human Confrontations." They of
course mean the power-law, which enjoys a special place in the
quantitative study of war. As some of the examples discussed in this
section highlight, scholars rely on the power-law for theorization of
war's microfoundations, prediction and missing value imputation, and
hypothesis testing. With such important objectives predicated on the
power-law, the choice to use it as opposed to other models has
far-reaching implications.

The idea that war intensity obeys the power-law---a simple mathematical
model that describes phenomena with many small magnitude events and few
extreme magnitude ones---originates with the influential physicist,
meteorologist, and quantitative scholar of human violence, Lewis F.
Richardson [-@richardson1948; -@richardson1960]. Sometimes referred to
as "Richardson's Law" [@clausetEtAl2007], the power-law of war holds a
privileged position as one of the few regularities in the social
sciences granted law-like status.

Tradition and convenience lie behind the inertia that propelled the
power-law to ubiquity. Power-law distributions have the unique (and
convenient) characteristic of linearity when plotted in log-log space.
The slope of the probability density distribution of a power-law over
the range of possible intensity levels of some variable in log-log space
lets one seamlessly map possible intensities to a specific probability
of occurrence. In addition to making projections of intensity
exceptionally easy, the method of estimating the relevant slope
parameter has a simple closed-form solution [@clausetEtAl2009].

Power-laws also have a rhetorically useful feature. Power-laws have only
a single parameter, and if this parameter takes a value at or less than
2 the mean of the variable in question can possibly be infinite. An
infinite mean implies that there is no upper limit on intensity (at
least none that is mathematically identifiable), making random variation
in war fatalities all the more alarming. Analyses such as Braumoeller's
[@braumoeller2019] have identified power-law slopes that are consistent
with this posibility. This gives shockingly ferocious and deadly
international wars the qualities of other "black swan" events, namely, a
high degree of unpredictability coupled with massive impact
[@taleb2010].

With few exceptions [see @zwetsloot2018], the power-law abides as the
framework of choice for studying war. With respect to theory generation,
@cederman2003 predicated an agent-based model of war on the stylized
fact that the most intense wars seem to conform to the power-law. His
study establishes microfoundations of human conflict that generate
power-law distributed war intensities.

Scholars have extended the power-law to other dimensions of conflict,
too. @cioffirevillaEtAl2010 conducted the first systematic replication
of Richardson's [-@richardson1948; @richardson1960] analysis, using the
most up-to-date COW data on international conflicts and extending the
analysis to civil wars as well. They also fit war onset and duration to
the power-law, expanding the range of war's characteristics that
plausibly conform to the power-law distribution. They conclude with a
call for more rigorous theorization of the data-generating processes
that give rise to these unique characteristics of conflict. Other
studies have used the power-law to study terrorism and all manner of
human violence, including attacks by lone-wolf actors [@clausetEtAl2007;
and @johnsonEtAl2013].

The power-law also serves as a framework for missing data imputation. In
one prominent example, @friedman2014 leveraged the power-law to make
inferences about unobserved fatalities in Native American and US
conflicts from 1776 to 1890. @friedman2014 lauds the potential of the
power-law to assist in missing value imputation for a range of important
conflicts for which data either are unavailable or untrustworthy.

Lastly, the power-law is used in hypothesis testing. @cedermanEtAl2011
collected data on the fatalities in major power wars since 1495 and fit
the data to the power-law using different successive splits in the
sample to identify a significant shift in war deadliness following the
Napoleonic Wars. @braumoeller2019, as already discussed, used the
power-law to test a claim from the decline-of-war thesis that wars
became less severe after 1945.

In sum, a lot rides on the appropriateness of the fit between observed
war intensity and the power-law; not just inferences important for the
decline-of-war thesis, but theorization, hypothesis testing, and
prediction more generally. A substantial mismatch between the data and
the power-law therefore would raise questions about the reliability of
analyses that use the approach. This would call for a reexamination of
several important studies (including many of those cited here).

## Possible Problems with the Power-Law

The power-law reigns supreme, in part, because of a preoccupation with
modeling the extreme ends of the distribution of war intensity. This
preoccupation comes as no surprise. However, the focus on the thick tail
of the distribution of wars has led to the neglect of the other end of
the distribution.

Methodology may bear some of the blame. @clausetEtAl2009 established the
state-of-the-art in fitting power-laws to empirical data. They provide a
helpful and more consistent closed-form maximum likelihood estimator for
parameter estimation in lieu of the more typical ordinary least squares
approach. They also recommend and describe a procedure for fitting a
power-law model to data that involves identifying a lower bound beyond
which the data conform well to the power-law, but below which the data
do not. This particular recommendation makes it practically feasible to
ignore smaller magnitude events in favor of events in the extreme tail
of the distribution.

Identifying a lower-bound has a practical purpose. Rarely do observed
data perfectly conform to our theoretical models of their generation,
and the need to identify a lower bound reflects this reality. But at
what point does this necessary concession to make the model fit data
slip into an effort to make the data fit the model? As @clausetEtAl2009
warn, their method will "tell us only the best fit to the power-law
form, not whether the power law is in fact a good model for the data"
(667).

The idea that there might be some inconsistencies between the power-law
and forms of human conflict has precedent. The most recent and
comprehensive critique to this author's knowledge appears in a working
paper by @zwetsloot2018. @zwetsloot2018 engages in a mammoth data
collection effort, assessing the fit of the power-law to over 685,000
conflict events across 16 datasets. His data is restricted to civil
conflict events for which data are publicly available, limiting our
ability to draw inferences from his analysis to major international
wars. Nonetheless, his analysis reveals a less-than-consistent fit
between the power-law distribution and a large number of civil conflict
events. Zwetsloot finds that the power-law provides an excellent fit for
only a third of the data and a probable fit for only half.

Recent studies raise questions about the appropriateness of the
power-law for a host of other phenomena from the physical and social
world as well. Connections within networks of various kinds have been
long thought to obey the power-law, but recent research that examines
more than 1,000 real-world networks from social, biological,
technological, and informational domains finds otherwise
[@broidoClauset2019]. Solar flares have also been argued to obey the
power-law, but a recent study suggests poor data bear the blame for this
inference. @verbeeckEtAl2019 find that after correcting for background
noise, the log-normal distribution better describes both peak flux and
fluence of solar flares, overturning decades of precedent.

In sum, the idea that the power-law provides the best description of war
intensity, while ubiquitous, deserves reexamination. The following
section discusses the technical details of the power-law model and
contrasts it with a proposed alternative: a log-logistic model of war
intensity. The latter model has the potential to better fit not only the
observations in the extreme tail of the distribution but also the less
extreme (and more likely) war intensities on the left side of the
distribution that fitting the power-law to the data requires we ignore.

## The Power-Law versus the Log-Logistic Distribution

Many probability distributions can characterize variables with many
small events and a few larger ones. These include exponential,
log-normal, or log-logistic distributions in addition to the power-law.
To narrow the scope of this study, in this section and those that follow
the power-law is contrasted with the log-logistic distribution. The
log-logistic model represents a promising alternative to the power-law
for characterizing war intensity for two reasons: (1) it should provide
a better fit for smaller wars while not interfering with the fit to more
extreme conflicts and (2) it has thicker tails than distributions such
as the log-normal which provides some additional advantages in dealing
with the thicker-tailed empirical distribution that various forms of
conflict tend to follow.

First, consider the power-law model of war deaths. The power-law
specifies that the probability of observing a war deadlier than some
value $x > 0$ is $$
\Pr(X > x) = \frac{\exp(\alpha)} {x^{\beta}} \quad \forall \quad x \geqslant x_{min} \tag{1}
$$ where $x_{min}$ is a lower bound such that $x$ follows a power-law
distribution to the degree $\beta > 0$ and $\alpha$ is a constant.
$\Pr(X >x)$ is one minus the empirical cumulative density of observed
war intensities. Expressed in log-log space, this is $$
\log[\Pr(X > 0)] = \alpha - \beta \log(x) \quad \forall \quad x \geqslant x_{min} \tag{2}
$$Given this convenient form, many studies have used ordinary least
squares to fit the model parameters. However, @clausetEtAl2009 show that
this approach can be inconsistent and recommend instead using maximum
likelihood. The maximum likelihood estimator for the slope $\beta$ has
the closed form solution: $$
\hat \beta = 1 + n / \sum_{i = 1}^n \log(x_i/x_{min}), \tag{3}
$$where $n > 0$ is the number of observations such that
$x \geqslant x_{min}$ where $i$ denotes the first element in this
vector.

While the solution for $\beta$ is simple enough, its appropriateness
depends on identifying the best $x_{min}$. Calculating $x_{min}$
requires a more involved process. @clausetEtAl2009 recommend choosing
different values of $x$ to serve as $x_{min}$ and estimating a new
$\beta$ per each iteration. Then, for each iteration, a
Kolmogorov-Smirnov test should be used to assess the fit between the
model and the data. The $x_{min}$ that provides the best fit is
selected.

As mentioned earlier, the need to identify a lower-bound is a necessary
and even justifiable concession in the face of real-world data. However,
if the lower-bound leads to a more than trivial truncation of
observations, the lack of perfect linearity between all $x$ and
$\Pr(X > x)$ in log-log space may actually signal poor specification. In
this case, it is worth considering whether other functional forms make
for a better fit for *all* the data---not just those beyond a lower
bound.

Unlike equation (1), the log-logistic model is specified as$$
\text{Odds}(X > x) = \frac{\Pr(X > x)}{1 - \Pr(X > x)} = \frac{\exp(\alpha)}{x^\beta}. \tag{4}
$$This can be expressed linearly as$$
\log[\text{Odds}(X > x)] = \alpha - \beta \log(x). \tag{5}
$$The idea with this specification is that the left-hand side of the
equation is better modeled as the log of the *odds* of observing a war
deadlier than $x$ than as the log of the *probability*. With this form,
the solutions for $\alpha$ and $\beta$ can be identified using the logit
maximum likelihood estimator.

Following @braumoeller2019, a bootstrapping procedure is used for
statistical inference for both the power-law and log-logistic
distributions.

In the sections that follow, the data used to compare the power-law and
log-logistic fits is presented. A pooled analysis that pits these two
probability distributions head to head follows. After this, I turn to a
replication and extension of @braumoeller2019 to see whether the
log-logistic fit provides different inferences relevant to the
decline-of-war thesis to those provided by the power-law.

## Data

To compare the log-logistic and power-law specifications for war
intensity I use a measure of total battle related deaths per capita of
major international wars from 1816 to 2007. I accessed the data using
the newly available `{peacesciencer}` R package which provides users
access to a wide range of datasets relevant to the study of conflict
[@peacesciencer-package]. Data on war deaths originated from the
Correlates of War Project's 4.0 data [@sarkeeswayman2010rw]. To ensure
that the measure reflects the relative intensity of wars, total battle
deaths were divided by the summed population of the countries involved
in fighting a given war. Estimates of country population data originated
from @andersetal2020bbgb.

The raw data was originally at the level of directed dyad-years. Before
analysis, the data were collapsed to the level of individual wars so
that each unit of observation was a unique war and the variable of
interest the total battle deaths per the populations of the countries
involved in the war. After aggregation, there were a total of 95 unique
observations. Total battle related deaths ranged from 2,000 to more than
134 million, with the median number of fatalities clocking in at just
over 20 thousand.

## Results

This section summarizes the results from a pooled comparison of the
power-law and log-logistic fits for war intensity. Figure 1 shows a
scatter plot of the probability of observing a war as deadly as that
observed in both log-log space and in logit-log space. Both the left and
right panels show the log of per capital battle deaths along the x-axis.
In the left panel the y-axis shows the log of the empirical
*probability* of a war deadlier than the one observed, and in the right
panel the y-axis shows the log of the empirical *odds* of observing a
war deadlier than the one observed. Each panel includes the estimated
model fit for the power-law and logit specifications, respectively. In
the case of the power-law fit, the regression line has slope
$\beta_{pl} = 1.66$ for observations past the lower bound $x_{min}$. For
the log-logistic fit, the regression slope is $\beta_{lg} = -0.74$.[^1]
Table 1 provides a summary of the estimates along with their
bootstrapped credible intervals.

[^1]: The power-law parameter is positive and the logit parameter
    negative simply due to the method of estimation. The linear
    relationships they correspond with in log-log space are both
    negative since the likelihood of ever deadlier conflicts is
    decreasing in war intensity.

```{=tex}
\begin{table}

\caption{Power-Law and Log-Logistic Parameter Estimates}
\centering
\begin{tabular}[t]{lcc}
\hline
 & \textbf{log[Pr(X > x)]} & \textbf{log[Odds(X > x)]}\\
\cline{2-3}
$x_{min}$ (\% War Deaths) & 0.025 [0.002; 0.156] & \\
Power-law Slope ($\beta_{pl}$) & 1.657 [1.408; 2.097] & \\
Log-logistic Intercept ($\alpha_{ll}$) & & -6.452 [-7.571; -5.566]\\
Log-logistic Slope ($\beta_{ll}$) & &-0.739 [-0.850; -0.636]\\
\hline
$N$ & 47 & 95 \\
\hline
\multicolumn{3}{l}{\textsuperscript{} Inference done with 200 bootstrapped samples. Bootstrapped 95\% confidence}\\
\multicolumn{3}{l}{\textsuperscript{} intervals shown in [brackets]. Estimates used to draw regression lines in Figure 1.}\\
\end{tabular}
\end{table}
```
At face value, the power-law specification provides a good fit for the
data past the lower bound. But, the loss of war deaths data with the
power-law fit is quite substantial with only less than 50% of all wars
providing the information necessary to model the probability of ever
deadlier conflicts. As evident from Figure 1, the power-law model is not
equipped to explain or predict the likelihood of war deaths in the the
remaining lower half of the sample.

While the power-law model poorly fits half the sample, the log-logistic
model provides an excellent fit for the *entire* sample. Across the
range of observed war deaths, the odds of observing a war deadlier than
the one observed appears to be well approximated by a linear function of
the observed magnitude of war deaths in log-log space. In fact, most
researchers would be fortunate to observe a linear fit a fraction as
tight as the one observed in Figure 1.

```{r, fig.cap="Empirical probability of wars as deadly as those observed in log-log space (left) and logit-log space (right). The latter provides a better fit for all the data while the former requires ignoring observations below a certain threshold. Table 1 estimates were used to draw the regression lines."}
include_graphics(
  here("02_figs", "fig1.png")
)
```

The limitations of the power-law model become all the more evident with
Figure 2. The x-axis shows the log of war deaths per 100 million, and
the y-axis shows the percent error in model fit relative to the data.
The solid line shows the error with the power-law model, and the dashed
line shows the error with the logit model. Positive values mean that the
model *overestimates* the probability of a war as deadly as the one
observed, while negative values mean that the model *underestimates* the
probability of a war as deadly as the one observed. Predictions are
based on the fitted parameters in Table 1, and the percent error is
calculated as the difference between the fit and the empirical CDF per
the value of the empirical CDF.

The difference in model fit for smaller wars shines through. The
power-law model poorly predicts the probability of war deaths around and
below 100 thousand per 100 million in population size. These are quite
deadly conflicts, making the poor fit of the power-law model for these
observations all-the-more consequential. Compare these poor predictions
with the far superior ones of the log-logistic model. While both models
make for nearly identical (and good) fits for the data for wars that
kill beyond 100 thousand per 100 million in population size, the
log-logistic model provides an exceptionally better fit for smaller wars
as well.[^2]

[^2]: The logistic fit is computed as
    $\hat{\Pr}(X > x) = F(\hat\alpha_{ll} + \hat \beta_{ll}\log[x])$,
    where $F(\cdot)$ is the logistic function. The power-law fit is
    computed as
    $\hat \Pr(X > x) = p \times (x / \hat x_{min})^{-\hat\alpha + 1}$,
    where $p$ is the observed $\Pr(X > \hat x_{min})$.

```{r, fig.cap="The percent error in model fit relative to the data. Model fit was obtained using the estimated parameters in Table 1. The percent error per a given conflict size was calculated as the difference in model fit relative to empirical CDF per the empirical CDF."}
include_graphics(
  here("02_figs", "fig2.png")
)
```

The limitations of the power-law fit go beyond prediction error for
smaller conflicts. While the log-logistic and power-law models yield
similar predictions for war deaths beyond the fitted power-law's lower
bound, the power-law model does so with far less precision.

Figure 3 shows the point-estimate of the power-law and logit model
predictions of the likelihood of observing a war that kills more than
10% of the belligerents' populations---a truly deadly conflict. The
x-axis shows the probability in percentages and the y-axis denotes the
respective model that generated the prediction. The 2.5 and 97.5
percentiles of the bootstrapped empirical sampling distribution of
predictions are included.

The point predictions generated by each model are nearly identical. Each
says that the likelihood of a war that kills more than 10% of the
populations of the countries fighting a war is just less than 1%.
However, while nearly identical, the logit model's predictions provide
more than double the precision than the power-law's. While the upper
bound on the bootstrapped 95% confidence interval for the power-law
prediction is just below 5%, the upper bound for the logit prediction is
well under 2%. These point estimates were obtained using the estimated
parameters in Table 1.

```{r, fig.cap="Comparison of power-law and logit model predictions of the probability that wars kill more than 10 percent of belligerent country populations. Predictions made using estimated parameters in Table 1."}
include_graphics(
  here("02_figs", "fig3.png")
)
```

## Implications: The Decline-of-War Revisited

The above analysis shows that the power-law provides a fairly poor fit
for war fatality data while the log-logistic distribution provides a far
superior fit. Does modeling war intensity via a log-logistic
distribution rather than the power-law yield different conclusions than
those arrived at in previous studies on war's deadliness?

As noted earlier, this question has special relevance in light of recent
debates about the so-called decline-of-war thesis. In response to
proponents of the theory, @pinker2011 being the most prominent target,
@braumoeller2019 makes short shrift of the decline-of-war thesis using,
among many statistical tests, the power-law to estimate changes in war
intensity over time. After devising a novel bootstrapping procedure for
testing whether two samples originate from different power-law
distributions, @braumoeller2019 applied the method to test whether a
detectable decline in war's deadliness occurred post-1945. Many
adherents to the decline-of-war thesis point to this date as a turning
point in war's severity, and a rejection of the null would support one
of their central claims, namely, that wars became less deadly in the
second half of the twentieth century. However, @braumoeller2019 failed
to identify a significant difference in war intensity. Even more, the
power-law estimates he identified suggest that war's intensity could
even have an infinite mean in both periods.

To test whether this finding holds up to an alternative model of war
intensity, this section summarizes the results from a replication of
Braumoeller's analysis followed by an extension that uses the
log-logistic fit for war intensity instead. The sample consisted of the
same data used in the previous analysis, save that it was divided into
two sub-samples denoting pre- and post-1945 conflicts. The first sample
consisted of 57 international conflicts, and the second of 38.

Using the same procedures for estimating the power-law and log-logistic
parameters outlined previously, it is possible to recover estimates of
the expected value of war fatalities for the different samples. For the
power-law distribution, the mean is identified as
$x_{min} \times (\beta - 1)/ (\beta - 2)$ where, if it is not already
plain to see, the mean is infinity if $\beta = 2$.[^3] Further, if
$\beta < 2$ the expected value of war deaths can also be negative which
is out of bounds for valid war fatalities. This means that only $\beta$
values less than 1 or greater than 2 are consistent with a definite and
non-negative mean for war deaths.

[^3]: See @clausetEtAl2009.

For the log-logistic distribution, the mean can simply be recovered from
the estimated slope and intercept parameters fit with the logit maximum
likelihood estimator. The identities of these parameters are in fact
defined in terms of the structural parameters $\mu$ and $\sigma$, where
the first is the mean of the log-logistic distribution and the second is
its standard deviation. The parameters $\mu$ and $\sigma$ can be easily
backed-out of the logit estimates for $\alpha$ and $\beta$ with the
identities of each given as:

$$
\alpha := \mu / \sigma \quad \text{and} \quad \beta := -1 / \sigma \tag{6}
$$

That implies that the relevant structural parameters can be calculated
as follows after estimation:

$$
\hat\sigma := -1 / \hat\beta \quad \text{and} \quad \hat\mu = \hat \alpha / \hat \beta. \tag{7}
$$

Table 2 provides a summary of the sample used in estimation. Power-law
and log-logistic models were fit to the data both pre- and post-1945.
The $N$ column denotes the total sample size for each period ($N = 57$
pre and $N = 38$ post). The column next to it indicates the effective
sample used to fit the power-law after identifying the best $x_{min}$.
The fact that the power-law leaves a good deal of the variation in war
intensity unexplained is made apparent by looking at the difference
between the sample sizes in the two columns. Only 34 out of 57 (60%)
wars provide a good fit for the power-law pre-1945, and only 13 out of
38 (34%) provide a good fit for the power-law post-1945.

```{=tex}
\begin{table}

\caption{Sample sizes pre- and post-1945}
\centering
\begin{tabular}[t]{lcc}
\hline
 & Total $N$ & Effective $N$ for Power-law\\
\cline{2-3}
post-1945 & 38 & 13\\
pre-1945 & 57 & 34\\
\hline
\end{tabular}
\end{table}
```
Figure 4 has two panels, each of which shows a scatter plot of the
empirical $\Pr(X > x)$ for war intensity on the y-axis and the observed
battle deaths per capita among the belligerents in a given war on the
x-axis. Values are shown in log-log space and battle deaths have been
converted to a rate per 100 million to provide a better sense of scale.
Color is used to differentiate pre-1945 (yellow) and post-1945 (blue)
conflicts. The left panel includes the line-of-best fit according to the
power-law model of war intensity. The right panel includes the
line-of-best fit according to the log-logistic model of war intensity.
Table 3 reports the parameter estimates, along with their bootstrapped
95% confidence intervals, used to plot the fits in the figure. On their
face, the results presented in Figure 4 demonstrate that the
log-logistic distribution makes for a much better fit for the data yet
again. While the power-law only applies to a truncated set of the
sample, the log-logistic model fits the entire sample (both pre- and
post-1945).

```{=tex}
\begin{table}

\caption{Paremeter estimates for log-logistic and power-law models}
\centering
\begin{tabular}[t]{lcc}
\hline
 & \textit{post-1945} & \textit{pre-1945}\\
\hline
\textbf{Log-logistic} & &\\
\cline{2-3}
$\quad\exp(\mu)$ & 6,349 [2,897; 11,343] & 27,469 [14,572; 46,412]\\
$\quad\sigma$ & 1.31 [1.02; 1.64] & 1.27 [1.04; 1.54]\\
\textbf{Power-law} & & \\
\cline{2-3}
$\quad x_{min}$ & 2,5471 [855; 45,044] & 24,680 [2,960; 210,229]\\
$\quad\beta$ & 1.81 [1.35; 2.52] & 1.62 [1.36; 2.54]\\
\hline
\multicolumn{3}{l}{\textsuperscript{} Inference done with 200 bootstrapped samples.}\\
\end{tabular}
\end{table}
```
```{r, fig.cap="A comparison of the power-law and log-logistic model fits for the data. The power-law fit is on the left and the log-logistic on the right. The data are shown in log-log space. War fatalities have been rescaled to battle deaths per 100 million."}
include_graphics(
  here("02_figs", "slopes.png")
)
```

Not only does the log-logistic distribution make for a better fit to the
data, statistical inference supports different conclusions about the
deadliness of wars after 1945. An examination of the parameters in Table
3 makes this clear. For the more visually inclined, Figures 5 and 6 plot
these estimates with their 95% confidence intervals.

Figure 5 shows the expected war deaths per 100 million both pre- and
post-1945 as suggested by the log-logistic distribution. The difference
is stark. According to the log-logistic fit for the data, the central
tendency for war intensity pre-1945 is more than 27,000 battle deaths
per 100 million. Conversely, the central tendency for war intensity
post-1945 is 6,394 per 100 million. This suggests wars after 1945 tend
to be orders of magnitude less deadly than those before 1945.

```{r, fig.cap="The expected level of war deadliness before and after 1945. Estimates obtained from a log-logistic specification of war intensity. See Tables 1 and 2 for sample and parameter estimates."}
include_graphics(
  here("02_figs", "ll_fig_45.png")
)
```

Contrast these conclusions with those implied by the slopes for the
power-law fit for the data. Figure 6 shows $\beta$ estimates for the
pre- and post-1945 samples. The slope estimates are shown rather than
the power-law mean because the range of possible estimates contain both
negative values and infinity. The estimated slopes for both periods are
between 1 and 2, implying a non-positive mean for war intensity, and the
confidence intervals for both periods overlap with 2, implying a
possibly infinite mean. While the estimated slope for the second period
is slightly larger (suggesting wars are *less* intense after 1945), the
slopes do not appear to be statistically distinguishable.

```{r, fig.cap="Power-law slopes for war deadliness before and after 1945. Estimates obtained from a power-law specification of war intensity. See Tables 1 and 2 for sample and parameter estimates."}
include_graphics(
  here("02_figs", "pl_fig_45.png")
)
```

Table 4 reports the results for bootstrapped tests of whether the
samples are drawn from different distributions according to the
respective models fit to the data. If we fit the power-law to the data,
there is insufficient evidence to reject the null that pre- and
post-1945 conflicts are drawn from different power-law distributions
($p = 0.86$). But while the null cannot be rejected in the case of the
power-law fits for the data, the null can be rejected that pre- and
post-1945 conflicts are drawn from the same log-logistic distribution
($p < 0.01$).

```{=tex}
\begin{table}

\caption{Difference in parameter estimates post- vs. pre-1945.}
\centering
\begin{tabular}[t]{lccc}
\hline
 & Estimate & 95\% Lower-Upper Bound & p-value \\
\cline{2-4}
$\Delta \mu_{ll}$ & -1.465 & [-2.527; -0.716] & 0.00\\
$\Delta \beta_{pl}$ & 0.188 & [-0.810; 0.964] & 0.94\\
\hline
\end{tabular}
\end{table}
```
## Discussion and Conclusion

These findings have important theoretical and practical implications for
the quantitative study of conflict. The idea that war deaths follow a
power-law distribution is the foundation for a range of work, from
theory generation to data imputation, and prediction to hypothesis
testing [@braumoeller2019; @cederman2003; @cedermanEtAl2011;
@cioffirevillaEtAl2010; @friedman2014]. Mispecification therefore
carries serious repercussions for the accumulation of scientific
knowledge of international conflict.

A replication and extension of recent analysis done by @braumoeller2019
makes this point clear. @braumoeller2019 set out to refute a claim from
the decline-of-war thesis that war's deadliness has declined over time,
particularly after 1945. To test this claim, he fit the power-law to
pre- and post-1945 data on international war intensities and devised a
statistical method to test the null that two samples come from the same
power-law distribution. Using this procedure, @braumoeller2019 failed to
reject the null using the 1945 cutoff---not evidence that this
particular claim of the the decline-of-war thesis is wrong, but an
indictment of the idea that the data provide strong support for it all
the same.

A replication of his procedure yields the same conclusion. However, when
extended to using the log-logistic fit for the data rather than the
power-law, the results change. A log-logistic fit suggests that
post-1945 conflicts have intensities significantly and orders of
magnitude less than pre-1945 conflicts.

The results from this study neither reflect a wholesale rebuttal to
@braumoeller2019, nor arrant support for the decline-of-war thesis. Many
factors other than those cited by decline-of-war proponents---like
Enlightenment Humanism or pacifistic norms---could explain the reduction
in conflict intensity after 1945. In fact, some clues may lie in
Braumoeller's [@braumoeller2019] discussion later in his book of the
role of international order in organizing patterns of international
conflict---a set of ideas that share some linkages with Lake's
characterization of international hierarchy [@lake2007; @lake2009].
Nonetheless, these results do raise questions about the reliability of
quantitative conflict research that relies on the power-law to the
exclusion of alternative probability distributions. Researchers may find
it worthwhile to revisit prior work, and in the future, scholars should
at minimum check the robustness of their inferences to alternative
methods. As far as the power-law and war go, perhaps the time has come
to lay this unqualified tradition to rest.

\clearpage

## References
