---
# title: "A Logistic Model of War Deaths is Better than the Power-Law"
# author: "Miles D. Williams"
# date: '2022-12-12'
bibliography: ../05_ref/ref.bib
editor_options: 
  markdown: 
    wrap: 72
fontfamily: mathpazo
fontsize: 12pt
geometry: margin=1in
header-includes:
- \usepackage{float}
- \usepackage{setspace}
- \usepackage{dcolumn}
- \doublespacing
- \setlength\parindent{24pt}
- \setlength\parskip{1pt}
- \usepackage[group-separator={,}]{siunitx}
- \usepackage{threeparttable}
- \usepackage{caption}\DeclareCaptionLabelFormat{AppendixTables}{Table A.#2}
- \usepackage{caption}\DeclareCaptionLabelFormat{AppendixFigures}{Figure A.#2}
- \usepackage[labelsep=period]{caption}
output:
  word_document:
    pandoc_args: 
      - '--lua-filter=parse-latex.lua'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      message = F,
                      warning = F,
                      fig.align = "center")
library(knitr)
library(here)
```

**Beyond the Power-Law of War: Why a Logistic Function of War's Deadliness is Better**

**Author**: Miles D. Williams^[Denison University, williamsmd@denison.edu.]

**Abstract**: The power-law specification of war deadliness is ubiquitous. In this study, an alternative logistic specification is proposed and shown to provide a superior fit for war fatality data and with greater statistical precision. Using Correlates of War data merged with estimates of country population totals, the relative intensity of major international wars fought between 1816 and 2007 are modeled using the power-law and logit. While the power-law only can be fit to data above a lower-bound of war intensity that requires dropping more than 50% of observations from the sample, the logistic model provides an excellent fit for all the data and with greater precision. Because the power-law specification of war intensity plays an essential role in theory generation, statistical inference, and hypothesis testing in the study of war, that an alternative specification exists that is both simple to estimate and that provides a superior fit should be welcome news.

**Keywords**: Conflict, Power-Law, Logit

**Word Count**: 4,145

\clearpage

## Introduction

In the social sciences, few regularities attain law-like status. The
idea that war deadliness follows a power-law distribution is one such
regularity, and scholars have regarded it as such since its initial
discovery earlier in the twentieth century [@richardson1948;
@richardson1960]. The practical, as well as alarming, significance of
war being power-law distributed has been addressed by numerous studies
over the past 10 to 20 years, but the idea received more widespread
popular attention than it had before with the publication of Bear
Braumoeller's [-@braumoeller2019] book *Only the Dead*.

The deadly potential of wars cannot be overstated, and the power-law
distribution applied to estimate the range of probable war fatalities
that belligerents suffer by war's end captures this well. However, the
wisdom of fitting war death data to the power-law rather than some other
distributional form is not always clear. A persistent practical reality
that arises in Braumoeller's [-@braumoeller2019] efforts to fit war
fatality data for all major wars fought over the last 200 years to the
power-law distribution is the fact that the power-law only provides a
good fit for the data above a lower bound. In many applied settings and
across a range of phenomena that plausibly follow a power-law
distribution, this fact is often treated as a minor hiccup that existing
statistical procedures have been designed to address [@clausetEtAl2009].
The solution entails identifying the lowest possible value of the
variable in question such that all the data at or above this threshold
conform to the power-law distribution. All data below this threshold are
then ignored.

The truncation of the data entailed by this solution is often
nontrivial. It is not impossible to lose more than 50% of the data
sample in an effort to make the data conform to the power-law
distribution.[^1] If such a large portion of the data must be ignored in
order to reliably fit the power-law to the data, it begs the question of
whether we are trying to make a square peg fit into a round hole. Is the
power-law really the best model of war intensity, or is there a better
alternative?

[^1]: Author's own analysis of Correlates of War (COW) data on war
    fatalities from 1816 to 2010.

The case is made here that the power-law *is not* the best fit for the
data and that there *is a better alternative*. Using data on the
intensity of major international wars from the past 200 years, a
comparison of the power-law model and a logistic model demonstrates the
superiority of the latter. The logistic model of war intensity not only
provides a much better fit for *all* of the data, the efficiency of the
logit model predictions is substantially higher than that of the
power-law model predictions in the extreme tails of the distribution of
war intensity. This finding has important implications for theory,
statistical inference, and hypothesis testing with war intensity
data[@braumoeller2019; @cederman2003; @friedman2014].

The paper proceeds as follows. Motivation and background on modeling
wars with the power-law is provided. This brief survey highlights the
near ubiquity of the idea that war intensity is power-law distributed.
With this ubiquity comes a number of implications for theory and
measures that are predicated on the power-law being the best fit for war
data.

A summary of the power-law model of war intensity and an alternative
logistic specification is then provided. A key difference in the two
specifications is the choice to model war intensity in log-log space as
probabilities versus odds. The latter may provide a better fit for the
seemingly shallow change in the likelihood of observing wars deadlier
than those observed among relatively low-intensity conflicts.

The data used to compare the power-law and logistic modelling approaches
is then summarized. Data on war fatalities are used from the Correlates
of War (COW) Project. War intensity is calculated as the total battle
deaths of all countries involved in a war per the combined population of
the relevant countries at the time of the conflict.

The results are then presented. Overall, the logistic model provides a
superior fit for the data in contrast to the power-law. The implications
are then discussed in the final section. While the power-law may apply
in numerous settings, in the case of international wars it provides a
poorer fit than many would suppose. In contrast to the power-law, the
logistic model offers a better fit and with greater statistical
precision. Future scholarship on war intensity would benefit from having
the logistic specification in their toolbox.

## Power-Laws and War Deaths

The idea that war intensity---the deadliness of wars relative to the
populations of the belligerent countries---is power-law distributed
received widespread attention with the recent publication of
Braumeoller's [-@braumoeller2019] *Only the Dead: The Persistence of War
in the Modern Age*. However, the "discovery" that war deadliness follows
the power-law is owed to Lewis F. Richardson [-@richardson1948;
-@richardson1960]. The power-law of war intensity enjoys a privileged
position as one of the few regularities in the social sciences granted
law-like status. This is a status that few have found cause to question
and for understandable reasons.

War fatalities, after all, have a highly skewed distribution. While
there are many wars in which few die, there are only a few wars in which
many die. This is just the sort of pattern that power-law distributed
data follow.

In addition, power-law distributions have the unique (and convenient)
characteristic of linearity when plotted in log-log space. The slope of
the probability density distribution of a power-law over the range of
possible intensity levels of some variable in log-log space lets one
seamlessly map possible intensities to a specific probability of
occurrence. In addition to making projections of intensity exceptionally
easy, the method of estimating the relevant slope parameter has a simple
closed-form solution.

Taken together, the fact of war intensity's skewed distribution and the
computational simplicity of the power-law make for a clear marriage of
convenience. But there is also a rhetorical motivation for modeling war
intensity with the power-law, upon which @braumoeller2019 capitalizes in
his conclusions about war's deadly potential. A feature of power-laws is
that if the slope parameter is at or less than 2 the mean of the
variable in question becomes infinity. An infinite mean implies that
there is no upper limit on intensity (at least none that is
mathematically identifiable), making random variation in war fatalities
all the more alarming. As it turns out, in Braumeoller's analysis, the
estimated power-law slope for war intensity is well below 2, implying
that war intensity can reach extremes far greater than even some of the
most volatile phenomena in the physical world such as fluid turbulence
or thermal spikes on the sun.

Braumoeller's is not the first attempt to model war intensity with the
power-law since @richardson1948. @cederman2003 predicated an agent-based
model of war on the stylized fact that the most intense wars conform to
the power-law. This stylized fact was supported by the same Correlates
of War (COW) data that Braumoeller would later use (albeit, an older
edition). @cioffirevillaEtAl2010 conducted the first systematic
replication of Richardson's[-@richardson1948; @richardson1960] analysis,
using the most up-to-date COW data on international conflicts and
extending the analysis to civil wars as well. They also fit war onset
and duration to the power-law, expanding the range of war's
characteristics that plausibly conform to the power-law distribution.

The ubiquity of war's power-law properties is not only a bit of trivia,
but it serves as a framework for inference and hypothesis testing.
@cedermanEtAl2011 collect data on the fatalities in major power wars
since 1495 and fit the data to the power-law using different successive
splits in the sample to identify a significant shift in war deadliness
following the Napoleonic Wars. @braumoeller2019 fits COW fatality data
to the power-law to test the claim from the "decline of war" theory that
conflicts have become less intense over time. Braumoeller develops a
statistical procedure for testing whether two power-law fits are
identified from different samples and finds that no significant change
in war's deadliness has occurred since 1816. Beyond inference for
hypothesis testing, @friedman2014 leverages the power-law to make
inferences about unobserved fatalities in Native American and US
conflicts from 1776 to 1890.

In sum, the power-law model of war intensity has several practical uses
for scholars. These include theory generation, statistical inference,
and hypothesis testing. A lot therefore rides on the appropriateness of
the fit between observed war intensity and the power-law, and a
substantial mismatch could raise questions about the reliability of
analyses that use the approach.

## A Power-Law or Logistic Model of War Intesnity?

The power-law reigns supreme, in part, because of the preoccupation with
modelling the extreme ends of the distribution of war intensity. War is
scary, but the idea that little but chance prevents any war from
escalating into the next world war (or worse) grips the imagination in a
visceral way. However, the focus on the thick tail of the distribution
of wars has led to the neglect of the other end of the distribution.
@braumoeller2019, for example, uses a procedure outlined by
@clausetEtAl2009 for fitting a power-law model to data that involves
identifying a lower bound beyond which the data conform well to the
power-law, but below which the data do not.

This procedure has a practical purpose. Rarely do observed data
perfectly conform to our theoretical models of their generation, and the
need to identify a lower bound reflects this reality. But at what point
does this necessary concession to make the model fit data slip into an
effort to make the data fit the model? As will be shown in the
subsequent analysis, the lower bound required to make the power-law fit
war intensity data leaves the intensity of more than half of the major
wars that occurred over the last 200 years unexplained. If nearly half
of the data fail to conform to a particular data-generating process, it
is worth asking whether a different model is more appropriate.

This section discusses the technical details of the power-law model and
contrasts it with a proposed alternative: a logistic model of war
intensity. This model has the potential to better fit not only the
observations in the extreme tail of the distribution but also the less
extreme (and thus more common) war intensities on the left side of the
distribution that fitting the power-law to the data requires we ignore.

First, consider the power-law model of war deaths. The power-law
specifies that the probability of observing a war deadlier than some
value $x > 0$ is $$
\Pr(X > x) = \frac{\exp(\alpha)} {x^{\beta}} \quad \forall \quad x \geqslant x_{min} \tag{1}
$$ where $x_{min}$ is a lower bound such that $x$ follows a power-law
distribution to the degree $\beta > 0$ and $\alpha$ is a constant.
Expressed in log-log space, this is $$
\log[\Pr(X > 0)] = \alpha - \beta \log(x) \quad \forall \quad x \geqslant x_{min} \tag{2}
$$A convenient feature of the power-law distribution is $\beta$ has a
closed-form solution. The maximum likelihood estimator for the slope
$\beta$ is estimated as: $$
\hat \beta = 1 + n / \sum_{i = 1}^n \log(x_i/x_{min}), \tag{3}
$$where $n > 0$ is the number of observations such that
$x \geqslant x_{min}$ where $i$ denotes the first element in this
vector.

While the solution for $\beta$ is simple enough, its appropriateness
depends on identifying the best $x_{min}$. Calculating $x_{min}$
requires a more involved process. We start by choosing different values
in $x$ to serve as $x_{min}$ and estimating a new $\beta$ per each
iteration. Then, for each iteration, we use a Kolmogorov-Smirnov test to
assess the fit between the model and the data. The $x_{min}$ that
provides the best fit is then selected.[^2]

[^2]: See @clausetEtAl2009 for a summary of the procedure.

This approach is applied by @braumoeller2019 and others; however, the
idea that there is a minimum threshold beyond which the power-law
distribution obtains is predicated on the idea that the power-law really
is the most appropriate model of the data. But the lack of perfect
linearity between all $x$ and $\Pr(X > x)$ in log-log space may actually
be a sign of poor specification. Other functional forms may make for a
better fit for *all* the data---not just those beyond a lower bound.

The logistic model is one such alternative that may be ideally suited to
handle the shallow-sloped distribution of war deaths below $x_{min}$.
Unlike equation (1), the logit model is specified as$$
\text{Odds}(X > x) = \frac{\Pr(X > x)}{1 - \Pr(X > x)} = \frac{\exp(\alpha)}{x^\beta}. \tag{4}
$$This can be expressed linearly as$$
\log[\text{Odds}(X > x)] = \alpha - \beta \log(x). \tag{5}
$$The idea with the logit specification is that the left-hand side of
the equation is better modeled as the log of the *odds* of observing a
war deadlier than $x$ than as the log of the *probability*. With this
form, the solutions for the intercept $\alpha$ and $\beta$ can be
identified using the logit maximum likelihood estimator.

A nice feature of the logistic model is that the relevant slope
parameter can be interpreted as an elasticity. The parameter $\beta$
reflects a percent change in the odds of a war deadlier than the one
observed per a percent change in war intensity.

Importantly, while the solutions for the relevant parameters of the
power-law and logit models of war deadliness are quite easy, inference
cannot be performed directly in either case. A tidy asymptotically
derived solution for the variance in model parameters does not exist, so
we instead need to use bootstrapping. Previous studies failed to be
attentive to this fact, and @braumoeller2019 was the first to develop
the approach.

The need for bootstrapping follows from the simple fact that estimation
of the relevant parameters depends on the empirical CDF of war
intensities, which is where much of the uncertainty in estimation comes
from. Neither the closed-form power-law solution, nor the logit maximum
likelihood estimator, take this uncertainty into account. The logit
standard errors, for example, would be far too narrow and thus lack
appropriate coverage. By bootstrapping the empirical sampling
distribution, this problem can be easily avoided.

## Data

To compare the logistic and power-law specifications for war intensity I
use a measure of total battle related deaths per capita of major
international wars from 1816 to 2007. I accessed the data using the
newly available `{peacesciencer}` R package which provides users access
to a wide range of datasets relevant to the study of conflict
[@peacesciencer-package]. Data on war deaths originated from the
Correlates of War Project's 4.0 data [@sarkeeswayman2010rw]. To ensure
that the measure reflects the relative intensity of wars, total battle
deaths were divided by the summed population of the countries involved
in fighting a given war. Estimates of country population data originated
from @andersetal2020bbgb.

The raw data was originally at the level of directed dyad-years. Before
analysis, the data were collapsed to the level of individual wars so
that each unit of observation was a unique war and the variable of
interest the total battle deaths per the populations of the countries
involved in the war. After aggregation, there were a total of 95 unique
observations. Total battle related deaths ranged from 2,000 to more than
134 million, with the median number of fatalities clocking in at just
over 20 thousand.

## Results

The results from comparing the power-law and logistic fits for war
intensity provide good evidence to favor the logistic specification over
the power-law. While the thick tailed predictions of both the logit and
power-law estimators are quite similar, the logistic model of war also
provides a good fit for smaller wars which the power-law fit is not
equipped to explain. Even more, because the logistic model can leverage
more information in the data, the efficiency of the logit model
predictions is substantially higher than that of the power-law model.

Figure 1 shows a scatter plot of the probability of observing a war as
deadly as that observed in both log-log space and in logit-log space.
Both the left and right panels show the log of per capital battle deaths
along the x-axis. In the left panel the y-axis shows the log of the
*probability* of a war deadlier than the one observed, and in the right
panel the y-axis shows the log of the *odds* of observing a war deadlier
than the one observed. Each panel includes the estimated model fit for
the power-law and logit specifications, respectively. In the case of the
power-law fit, the regression line has slope $\beta_{pl} = 1.66$ for
observations past the lower bound $x_{min}$. For the logit fit, the
regression slope is $\beta_{lg} = -0.74$.[^3] Table 1 provides a summary
of the estimates along with their bootstrapped credible intervals.

[^3]: The power-law parameter is positive and the logit parameter
    negative simply due to the method of estimation. The linear
    relationships they correspond with in log-log space are both
    negative since the likelihood of ever deadlier conflicts is
    decreasing in war intensity.

```{=tex}
\begin{table}

\caption{Power-Law and Logistic Parameter Estimates}
\centering
\begin{tabular}[t]{lrrr}
\hline
  & Estimate & Lower & Upper\\
\hline
\multicolumn{4}{l}{\textbf{Power-Law}}\\
\hspace{1em}x-min (\% of population) & 0.025 & 0.002 & 0.156\\
\hspace{1em}Slope & 1.657 & 1.408 & 2.097\\
\multicolumn{4}{l}{\textbf{Logistic}}\\
\hspace{1em}Intercept & -6.452 & -7.571 & -5.566\\
\hspace{1em}Slope & -0.739 & -0.850 & -0.636\\
\hline
\multicolumn{4}{l}{\textsuperscript{} Inference done with 200 bootstrapped samples.}\\
\end{tabular}
\end{table}
```

At face value, the power-law specification provides a good fit for the
data past the lower bound. But, the loss of war deaths data with the
power-law fit is quite substantial with only less than 50% of all wars
providing the information necessary to model the probability of ever
deadlier conflicts. As is evident from Figure 1, the implication is that
the power-law model is not equipped to explain or predict the likelihood
of war deaths in the the remaining lower half of the sample.

While the power-law model poorly fits half the sample, the logit model
provides an excellent fit for the *entire* sample. Across the range of
observed war deaths, the odds of observing a war deadlier than the one
observed appears to be well approximated by a linear function of the
observed magnitude of war deaths in log-log space. In fact, most
researchers would be fortunate to observe a linear fit a fraction as
tight as the one observed in Figure 1.

```{r, fig.cap="Empirical probability of wars as deadly as those observed in log-log space (left) and logit-log space (right). The latter provides a better fit for all the data while the former requires ignoring observations below a certain threshold."}
include_graphics(
  here("02_figs", "fig1.png")
)
```

The problem of data fit with the power-law model becomes all the more
evident with Figure 2. The x-axis shows the log of war deaths per 100
million, and the y-axis shows the percent error in model fit relative to
the data. The solid line shows the error with the power-law model, and
the dashed line shows the error with the logit model. Positive values
mean that the model *overestimates* the probability of a war as deadly
as the one observed, while negative values mean that the model
*underestimates* the probability of a war as deadly as the one observed.

The difference in model fit for smaller wars shines through. The
power-law model is poorly suited to predicting the probability of war
deaths around and below 100 thousand per 100 million in population size.
These are quite deadly conflicts, making the poor fit of the power-law
model for these observations all-the-more consequential. Compare these
poor predictions with the far superior ones of the logit model. While
both models make for nearly identical (and good) fits for the data for
wars that kill beyond 100 thousand per 100 million in population size,
the logit model provides an exceptionally better fit for smaller wars as
well.

```{r, fig.cap="The percent error in model fit relative to the data."}
include_graphics(
  here("02_figs", "fig2.png")
)
```

The limitations of the power-law fit go beyond prediction error for
smaller conflicts. While the logit and power-law models yield similar
predictions for war deaths beyond the fitted power-law's lower bound,
the power-law model only does so with less precision than the logit
model. Further, the reduced precision of the power-law fit is far from
trivial.

Figure 3 shows the point-estimate of the power-law and logit model
predictions of the likelihood of observing a war that kills more than
10% of the belligerent country populations---a truly deadly conflict.
The x-axis shows the probability in percentages and the y-axis denotes
the respective model that generated the prediction. The 2.5 and 97.5
percentiles of the bootstrapped empirical sampling distribution of
predictions are included.

The point predictions generated by each model are nearly identical. Each
says that the likelihood of a war that kills more than 10 percent of the
populations of the countries fighting a war is just less than 1%.
However, while the predictions are nearly identical, the logit model
provides more than double the precision than the power-law model. While
the upper bound on the bootstrapped 95% confidence interval for the
power-law prediction is just below 5%, the upper bound for the logit
prediction is well under 2%.

```{r, fig.cap="Comparison of power-law and logit model predictions of the probability that wars kill more than 10 percent of belligerent country populations."}
include_graphics(
  here("02_figs", "fig3.png")
)
```

## Discussion and Conclusion

The idea that war's deadliness is power-law distributed has gained
wide-spread acceptance. Yet, the power-law is not the only possible
data-generating process that can be used to model war intensity. As this
study shows, it may not even be the best. When compared head to head,
the logistic model of war intensity performs much better than the
power-law. It provides a solid fit for the entire distribution of
intensities of international wars that have been fought for the past 200
years, not just the most extreme 50%. Further, the logistic model
provides nearly identical probability estimates of varying war
intensities to the power-law model in the extreme tails of the
distribution, but it does so with more than twice the statistical
precision.

These findings have important theoretical and practical implications for
the study of war. The idea that war deaths follow a power-law
distribution is not merely a matter of trivia but also a basis for
theory generation, statistical inference, and hypothesis testing
[@braumoeller2019; @cederman2003; @cedermanEtAl2011;
@cioffirevillaEtAl2010; @friedman2014]. Mispecification is therefore not
just matter of getting the data-generating process of war intensity
right. It is also about providing the best framework for theorizing
about war's microfoundations, identifying whether significant shifts in
intensity have occurred over time, or whether we can reliably infer war
fatalities where we otherwise lack the objective data to do so.

The results from this study will hopefully ensure that future
scholarship does not proceed as though only one model of war intensity
will do. The power-law may certainly be a good fit for some phenomena,
and even for some dimensions of international conflict other than
intensity (like duration or onset). Analysis of the wider array of war's
characteristics is beyond the scope of this research note. Nonetheless,
it is important that scholars are aware that there are alternative
models of war characteristics, with the logistic model serving as one
excellent choice---to this author's knowledge, one that has not been
applied. There is no point trying to make a square peg fit into a round
hole if we know a round peg is within reach.

\clearpage

## References
